---
title: "Stroke"
author: "Pascal Visser"
date: "2022-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# EDA

## intro

A stroke is a life-threatening medical condition, where there is no or poor blood flow to (parts) the brain. This causes parts of the brain to stop functioning and over minutes time die off. A stroke or brain attack is labelled as the second leading cause of death worldwide and is responsible for an annual mortality of about 5.5 million.

This dataset consists of people that had a stroke and people that have not. The variables are values that indicate there live style and environment. The goal of this research is to calculate the chance of a stroke based of there lifestyle. This can be accomplished by machine learning. This gives the following research question:

*Is it possible to produce an accurate algorithm with machine learning, to calculate the risk of a stroke based on life style variables of people with stroke history and people that never had a stroke?* 



## Dataset

The data is as said, lifestyle information of people with stroke history or not. In this section the data will be talked trough and explaned.
First the data is loaded in:

```{r}
strokedata <- read.csv("Data/Stroke_dataset.csv", header = T)
head(strokedata)
```

```{r}
codebook <- read.table("Data/codebook.txt", header = T, sep = ';')
codebook
```


# program part

- Load dataset
- Modify dataset?
- Dataset statistics 


